---
title: "Pokemon Analysis"
author: "Connor Bryson, Nathen Byford, Miguel Iglesias"
date: "10/12/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,fig.height = 2.5)
library("patchwork")
library("tidyverse")
theme_set(
  theme_bw() +
    theme(
      panel.grid.minor.x = element_blank(), 
      axis.title.y = element_text(angle = 0, vjust = .5)
    )
)
```


```{r data cleaning, echo=FALSE, warning=FALSE, message=FALSE}
# read in data.
pokemon <- read_csv(".\\data\\pokemon.csv")

# make function to factorize happiness
factorize <- function(x) {
  factor(case_when(
            x == (0) ~ "Very unhappy",
            x == (35) ~ "Moderatly unhappy",
            x == (50) ~ "Slightly happy",
            x == (70) ~ "Neutral",
            x == (90) ~ "Slightly unhappy",
            x == (100) ~ "Moderatly happy",
            x == (140) ~ "Very happy"
            ), 
         levels = c(
           "Very unhappy", 
           "Moderatly unhappy", 
           "Slightly happy",
           "Neutral",
           "Slightly unhappy",
           "Moderatly happy",
           "Very happy"
           ), 
         ordered = TRUE
         )
}

# New pokemon dataset to use, adds BMI and happiness
pokemon_new <-  pokemon %>%
  separate(typing, into = c("primary", "secondary")) %>%
  mutate(BMI = (weight/10)/((height/10)^2)) %>%
  filter(BMI <= 50000) %>%
  mutate(happiness = factorize(factor(base_happiness, ordered = TRUE)))

```

## Introduction


<<<<<<< HEAD
=======


>>>>>>> a0d9fa52d203765799775d3b3883cd73722afec4
## Questions and Findings

### Speed

We want to begin with looking into what factors effect the speed of the Pokemon. 


### Capture Rate

We also wanted to look into how capture rate is effected by variables such as height, weight, gender, and legendary.


### Happiness

Lastly we wanted to see what factors effected the happiness of the Pokemon. This variable was just something we thought would be fun to look into. Our main questions were what other variables effect the Pokemon's happiness. We began with looking at if the Pokemon is a legendary and if it's a mythical Pokemon.

We were also curious if the female rate effected happiness of the Pokemon. Converting the happiness variable to a likert scale we can easily see the amount of data points in each bin of female rate in the following plot. As seen in the plots before there is a smaller amount of Pokemon with a high female rate. Notice how it appears that depending of the female rate there aredifferent levels of happiness that are observed. What stood out to us what that Pokemon with a female rate of 0.25 or 0.875 are only observed with neutral happiness and there are no very unhappy Pokemon with a rate of 1. There is no apparent correlation between happiness and female rate.

```{r}
pokemon_new |> ggplot(aes(as.factor(female_rate), happiness)) +
  geom_bin2d(alpha = .9) +
  scale_fill_viridis_c() +
  labs(title = "Happiness of Pokemon by Female Rate", x = "Female Rate", y = "Happiness") +
  theme(panel.grid.major.x = element_blank())
```

