---
title: "Pokemon Analysis"
author: "Connor Bryson, Nathen Byford, Miguel Iglesias"
date: "10/12/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,fig.height = 2.5)
library("patchwork")
library("ggmosaic")
library("tidyverse")
theme_set(
  theme_bw() +
    theme(
      panel.grid.minor.x = element_blank(), 
      axis.title.y = element_text(angle = 0, vjust = .5),
      plot.title = element_text(hjust = 0.5)
    )
)
```


```{r data cleaning, echo=FALSE, warning=FALSE, message=FALSE}
# read in data.
pokemon <- read_csv(".\\data\\pokemon.csv")

# make function to factorize happiness
factorize <- function(x) {
  factor(case_when(
            x == (0) ~ "Very unhappy",
            x == (35) ~ "Moderatly unhappy",
            x == (50) ~ "Slightly happy",
            x == (70) ~ "Neutral",
            x == (90) ~ "Slightly unhappy",
            x == (100) ~ "Moderatly happy",
            x == (140) ~ "Very happy"
            ), 
         levels = c(
           "Very unhappy", 
           "Moderatly unhappy", 
           "Slightly happy",
           "Neutral",
           "Slightly unhappy",
           "Moderatly happy",
           "Very happy"
           ), 
         ordered = TRUE
         )
}

# New pokemon dataset to use, adds BMI and happiness
pokemon_new <-  pokemon %>%
  separate(typing, into = c("primary", "secondary")) %>%
  mutate(BMI = (weight / 10) / ((height / 10)^2)) %>%
  filter(BMI <= 50000) %>%
  mutate(happiness = factorize(factor(base_happiness, ordered = TRUE)))

```

## Introduction

<<<<<<< HEAD
Pokémon, a Japanese card and video game, revolves around a fantasy world where people fight each other with creatures they find and domesticate. Each creature, or Pokémon, can be characterized by certain attributes from health points (hp), to their typing which includes fire-Pokémon, water-Pokémon, and grass-Pokémon among others. Our dataset is a CSV that takes all the defined characteristics of the Pokémon and collects into one useable file. Though the dataset includes 49 variables, the ones we are using include: ___________. 
=======
Pokémon, a Japanese card and video game, revolves around a fantasy world where people fight each other with creatures they find, capture, and then domesticate. Each creature, or Pokémon, can be characterized by certain attributes including type (fire, grass, poison etc.), identifiers (legendary vs non-legendary etc.), and skill points (health, attack, defense etc.). Our dataset is a CSV that takes all the attributes of every Pokémon and collects into one usable file. Though the dataset includes 49 variables, we primarily looked at speed, capture rate, and happiness. We choose these variables for a variety of reasons. Speed was chosen, because during the exploratory portion of the project we found that speed and weight had no significant correlation. So, I [Connor] decided to investigate if the developers of the game correlated speed to any other variables. The lack of relationship between weight and speed made us distrust the development of the game. It seemed as though the developers did not put much thought into the values they gave each Pokémon. Given that we were already looking at speed, I [Miguel] wanted to see if the capture rate had any predictors. I [Nate] decided to do the same but with the base happiness of the Pokémon. Each portion of the analysis will be discussed in more detail throughout the paper to give context to variables and analysis ideas.

![The pokedex entry for Pikachu. Includes info such as type, height, and weight](".\\Pokedex_Picture")

>>>>>>> e14794bf5c0251c6de984ddef72b80e3f9e383e3

## Questions and Findings

### Speed

I [Connor] want to begin with looking into what factors effect the speed of the Pokemon.
This included testing whether real life factors such as weight and height affected the speed of the Pokemon. At first we believed they would, but as seen by the graphs below we were wrong. 

```{r}
#Weight vs Speed (organized by Generation)
spdvswgt <- ggplot(pokemon, aes(x =weight, y =speed, color= gen_introduced)) +
  geom_point()+
  labs(title="Weight vs Speed", x= "Weight", y="Speed")+
  xlim(c(0, max(pokemon$weight))) + 
  ylim(c(0, max(pokemon$speed)))+
  viridis::scale_color_viridis()
 

#Speed vs Height (organized by Generation)
spdvshgt <- ggplot(pokemon, aes(x =speed, y =height, color= gen_introduced)) +
  geom_point()+
  labs(title="Speed vs Height", x= "Speed", y="Height")+
  xlim(c(0, max(pokemon$speed))) + 
  ylim(c(0, max(pokemon$height)))+
  viridis::scale_color_viridis()+
  theme(plot.title = element_text(hjust = 0.5))

spdvswgt
spdvshgt
```
Looking at the graph for weight and speed, we noticed that there are outliers (on the right side of the graph) that do not make sense for speed. To see whether this was a mistake or not, we looked at the generation of each Pokemon  and found that the majority of the outliers came from the most recent generation rather than the earlier generations when Pokemon's system of attributes were still in the works. In addition to weight, height did not have an affect on the speed of a Pokemon, thus defying all of our expectations of this data set.


Afterwards, we wanted to see if the speed of a Pokemon affected its overall capture rate and whether it is harder to catch a Pokemon if it has a higher speed or not. 
```{r}
#Speed vs Capture Rate
ggplot(pokemon, aes(x =speed, y =capture_rate,  color= gen_introduced)) +
  geom_point()+
  labs(title="Speed vs Capture Rate", x= "Speed", y="Capture Rate")+
  xlim(c(0, max(pokemon$speed))) + 
  ylim(c(0, max(pokemon$capture_rate)))+
  geom_smooth(se=FALSE)+
  viridis::scale_color_viridis()+
  theme(plot.title = element_text(hjust = 0.5))

```
The plot shows a general negative trend as the speed increases, the capture rate decreases.


Finally with speed, since there was no realistic interpretation of speed for a given Pokemon, we decided to see if there was a relationship between a Pokemon's speed of attack and defense.  

```{r}
#Speed vs Attack 
spdvsatk <-ggplot(pokemon, aes(x =speed, y =attack,  color= gen_introduced)) +
  geom_point()+
  labs(title="Speed vs Attack", x= "Speed", y="Attack")+
  xlim(c(0, max(pokemon$speed))) + 
  ylim(c(0, max(pokemon$attack)))+
  geom_smooth()+
  viridis::scale_color_viridis()+
  theme(plot.title = element_text(hjust = 0.5))

#Speed vs Defense
spdvsdef <-ggplot(pokemon, aes(x =speed, y =defense,  color= gen_introduced)) +
  geom_point()+
  labs(title="Speed vs Defense", x= "Speed", y="Defense")+
  xlim(c(0, max(pokemon$speed))) + 
  ylim(c(0, max(pokemon$defense)))+
  geom_smooth()+
  viridis::scale_color_viridis()+
  theme(plot.title = element_text(hjust = 0.5))

spdvsatk+spdvsdef
```
The results found that attack tends to have a slight positive correlation with speed while defense does not. 



### Capture Rate

#### Size versus Capture Rate

As mentioned earlier, I [Miguel] wanted to test the game developers and see if anything predicted capture rate. Capture rate describes how easy it is to capture a Pokémon found in the wild. The lower
the capture rate, the harder the Pokémon is to capture. For more information on capture rate, a detailed description can be found at: https://bulbapedia.bulbagarden.net/wiki/Catch_rate. Of the variables in the dataset I first decided to look at the influence of height, weight, and then BMI. It made intuitive sense to me that bigger Pokémon would be harder to capture so I decided to test it out.

``` {r}


#Create a plot that shows the relationship between capture_rate and BMI
#Limit to not include a lot of empty space
BMI <- ggplot(pokemon_new, aes(BMI, capture_rate)) +
  geom_jitter(alpha = 0.2, color = "Steel Blue", size = .5) +
  geom_smooth(se = FALSE, color = "Steel blue") +
  coord_cartesian(ylim = c(0,275)) +
  labs(x = "Body Mass Index", y = "Capture Rate")


#Create a plot that shows the relationship between capture_rate and Height
#Limit included to allow for best view of relationship
Height <- ggplot(pokemon_new, aes(height/10, capture_rate)) +
  geom_jitter(alpha = 0.2, color = "Steel Blue", size = .5) +
  geom_smooth(se = FALSE, color = "Steel blue") +
  coord_cartesian(xlim = c(0,5), ylim = c(0,275)) +
  labs(x = "Height (m)", y = NULL)
  
#Create a plot that shows the relationship between capture_rate and weight
#Limit included to allow for best view of relationship

Weight <- ggplot(pokemon_new, aes(weight/10, capture_rate)) + 
  geom_jitter(alpha = 0.2, color = "Steel blue", size = .5) +
  geom_smooth(alpha = 0.5, se = FALSE , color = "Steel blue") +
  coord_cartesian(ylim = c(0,275)) +
  labs(x = "Weight (kg)", y = "Capture Rate")


(Weight | Height)
BMI

```

As can be seen, size does have an influence on the rate of capture. The taller and heavier Pokémon both have decreased rates of capture. Interestingly enough, however, when the two are put together as with BMI, the relationship goes away. To further understand the complex relationship between size and capture rate, more analysis would be required, however, the preliminary look suggests that size does have some affect on capture rate.
#### Other variables of interest

After learning more about the relationship between size and capture rate, I [Miguel] decided to see how other variables may affect the capture rate. The first one I examined was the shape of the Pokémon. Shapes include armor, quadraped, upright and more. 
```{r}

p1 <- ggplot(pokemon_new) +
  geom_boxplot(aes(reorder(shape, capture_rate, FUN = median), capture_rate)) +
  labs(x = "Shape", y = "Capture Rate")

p1
```

As seen in the plot above, shape has little influence on capture rate except for in the cases of Legs and Armor. What may be surprising, however, is that armored Pokémon are easier to capture than non-armored Pokémon.

Next, I looked at four of the boolean variables from the dataset. I wanted to see if how capture rate was influenced by Pokémon being Legendary, Mythical, genderless, and babies. One might expect mythical and legendary Pokémon to be harder to catch, babies easier to catch, and genderless Pokémon to have about equal rates to the gendered ones.

```

p2 <- ggplot(pokemon_new, aes(mythical, capture_rate)) + 
  geom_boxplot() +
  labs(x = "Mythical", y = "Capture Rate")

p3 <- ggplot(pokemon_new, aes(legendary, capture_rate)) + 
  geom_boxplot() +
  labs(x = "Legendary", y = NULL)

p4 <- ggplot(pokemon_new, aes(genderless, capture_rate)) + 
  geom_boxplot() +
  labs(x = "Genderless", y = NULL)

p5 <- ggplot(pokemon_new, aes(baby_pokemon, capture_rate)) + 
  geom_boxplot() +
  labs(x = "Baby Pokemon", y = NULL)

p2 | p3 | p4 |p5
  
```

As expected, legendary and mythical Pokémon are more difficult to catch, and babies are easier to catch. What is surprsing, however, is that genderless Pokémon are also harder to catch. This could be due to a confounding variable, however, it is more likely that the game developers did not put that much thought into this. In fact, a preliminary look for confounding variables suggested that the genderless variable is not confounded by the legendary or mythical variable, both of which might have explained the increased difficulty of capture for genderless Pokémon.

Lastly, I looked at how the percentage that each Pokémon is female influences capture rate. Within the game, Pokémon can be bred to make new Pokémon. Due to the presence of Pokémon breeding, Pokémon gender must also exist and each Pokémon has a set probability of being female when found in the wild. Some Pokémon have a female rate of 12.5%, whereas others are as high as 100%. This does not mean a Pokémon can be 12.5% female and 87.5% male, merely that within the whole population 12.5% of the Pokémon are female. Genderless Pokémon have a female rate of 0.
  
``` {r}
 
p6 <- ggplot(pokemon_new, aes(as.factor(female_rate*100), capture_rate)) +
  geom_boxplot() +
  labs(x = "Percent Female", y = "Capture Rate")

p6
 
```

To avoid any confusion, it must be understood that within each Pokémon, the males and females have each capture rate. Though, as it turns out, the female rate does influence the capture rate. Generally speaking, the Pokémon with high female rates are easier to capture. Though the 0% female category may be lower due to the addition of the genderless Pokémon, the 12.5% and 25% catergories are also low, especially compared to the 87.5% category. Why male-dominant Pokémon are harder to capture than female-dominant Pokémon is a question for another day.
 
### Happiness

Lastly we wanted to see what factors effected the happiness of the Pokemon. This variable was just something we thought would be fun to look into. Our main questions were what other variables effect the Pokemon's happiness. We began with looking at if the Pokemon is a legendary and if it's a mythical Pokemon.

To get a grasp of what effects different variables had on the happiness of the Pokemon we took a look at how different characteristics of the Pokemon had an effect. Bellow are three violin plots looking at mythical, legendary, and baby Pokemon and comparing them to Pokemon without that attribute based on happiness. It is  fairly obvious that mythical and legendary Pokemon tend to be less happy, I thinkk that could be that they are tougher pokemon so when they are captured they are not happy about it. These Pokemon like to have free reign and not be controlled by trainers. While the Baby Pokemon look to be as happy or more happy than most other pokemon.  

```{r, fig.height=2}
v1 <- pokemon_new |> ggplot(aes(mythical, base_happiness)) +
  geom_violin() +
  labs(title = "Mythical Pokemon", x = "Mythical", y = "Happiness")

v2 <- pokemon_new |> ggplot(aes(legendary, base_happiness)) +
  geom_violin() +
  labs(title = "Legendary", x = "Legendary", y = NULL)

v3 <- pokemon_new |> ggplot(aes(baby_pokemon, base_happiness)) +
  geom_violin() +
  labs(title = "Baby Pokemon", x = "Baby Pokemon", y = NULL)

(v1 + v2 + v3)
```

We also looked at how the Pokemon primary type effects it's happiness. Looking at the mosaic plot below we can see that there are different distributions of happiness bases upon primary type. Some observations to be made is that the most unhappy pokemon types are dragon, dark, and steel. Also the majority of Pokemon appear to be neutral.

```{r}
m1 <- pokemon_new |> ggplot() +
  geom_mosaic(aes(x = product(happiness, primary), fill = happiness)) + 
  labs(title = "Pokemon Happiness by Primary Type", x = "Primary Type", 
       y = "Happiness") +
  theme(
    panel.grid.major.y = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.text.x.bottom = element_text(angle = 60, vjust = 1, hjust = 1)
  )

m1
```


We were also curious if the female rate effected happiness of the Pokemon. Converting the happiness variable to a likert scale we can easily see the amount of data points in each bin of female rate in the following plot. As seen in the plots before there is a smaller amount of Pokemon with a high female rate. Notice how it appears that depending of the female rate there are different levels of happiness that are observed. What stood out to us what that Pokemon with a female rate of 0.25 or 0.875 are only observed with neutral happiness and there are no very unhappy Pokemon with a rate of 1. There is no apparent correlation between happiness and female rate.

```{r}
pokemon_new |> ggplot(aes(as.factor(female_rate), happiness)) +
  geom_bin2d(alpha = .9) +
  scale_fill_viridis_c() +
  labs(title = "Happiness of Pokemon by Female Rate", x = "Female Rate", y = "Happiness") +
  theme(panel.grid.major.x = element_blank())
```

