---
title: "Pokemon Analysis"
author: "Connor Bryson, Nathen Byford, Miguel Iglesias"
date: "10/12/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,fig.height = 3)
library("patchwork")
library("tidyverse")
theme_set(
  theme_bw() +
    theme(
      panel.grid.minor.x = element_blank(), 
      axis.title.y = element_text(angle = 0, vjust = .5)
    )
)
```


```{r data cleaning, echo=FALSE, warning=FALSE, message=FALSE}
# read in data.
pokemon <- read_csv(".\\data\\pokemon.csv")

# make function to factorize happiness
factorize <- function(x) {
  factor(case_when(
            x == (0) ~ "Very unhappy",
            x == (35) ~ "Moderatly unhappy",
            x == (50) ~ "Slightly happy",
            x == (70) ~ "Neutral",
            x == (90) ~ "Slightly unhappy",
            x == (100) ~ "Moderatly happy",
            x == (140) ~ "Very happy"
            ), 
         levels = c(
           "Very unhappy", 
           "Moderatly unhappy", 
           "Slightly happy",
           "Neutral",
           "Slightly unhappy",
           "Moderatly happy",
           "Very happy"
           ), 
         ordered = TRUE
         )
}

# New pokemon dataset to use, adds BMI and happiness
pokemon_new <-  pokemon %>%
  separate(typing, into = c("primary", "secondary")) %>%
  mutate(BMI = (weight/10)/((height/10)^2)) %>%
  filter(BMI <= 50000) %>%
  mutate(happiness = factorize(factor(base_happiness, ordered = TRUE)))

```

## Introduction




## Questions and Findings

### Speed

We want to begin with looking into what factors effect the speed of the Pokemon. 


### Capture Rate

We also wanted to look into how capture rate is effected by variables such as height, weight, gender, and legendary.


### Happiness

Lastly we wanted to see what factors effected the happiness of the Pokemon. This variable was just something we thought would be fun to look into. Our main questions were what other variables effect the Pokemon's happiness. We began with looking at if the Pokemon is a legendary and if it's a mythical Pokemon.
